% ══════════════════════════════════════════════════════════════════════
% End Matter: Acknowledgements, Funding, COI, Author Contributions,
%             AI Disclosure, Data Availability
% ══════════════════════════════════════════════════════════════════════

\section*{Acknowledgements}

The author thanks the open-source maintainers of Stim, PyMatching, and PyTorch for the infrastructure on which this work was built.

\section*{Funding}

This work received no external funding.

\section*{Competing Interests}

The author declares no competing interests.

\section*{Author Contributions}

Mohammad Albataineh directed the overall project workflow, selected and prioritized research directions, reviewed and validated experimental outputs, integrated the final manuscript and release artifacts, and approved the final version. Large language model tools were used extensively for coding assistance, debugging, drafting, editing, formatting, and document preparation.

\section*{AI Assistance Disclosure}

This work was developed through a heavily AI-assisted workflow. Large language model tools were used extensively across research planning support, experiment orchestration, code generation and revision, debugging, manuscript drafting and editing, and \LaTeX/PDF preparation. The author's role was to direct and manage the workflow, review generated material, validate the final outputs used in the paper, integrate the final manuscript and project artifacts, and assume full responsibility for the scientific claims, reported results, and interpretations.

\section*{Data and Code Availability}

The code, experiment artifacts, and reproduction scripts associated with this work are contained in the project repository and release bundle. Public access details will be provided at publication time.
