% ══════════════════════════════════════════════════════════════════════
% Section 6: Related Work
% ══════════════════════════════════════════════════════════════════════

\section{Related Work}

\paragraph{ML decoders for QEC.} Neural approaches to surface-code decoding include feedforward networks~\cite{PLACEHOLDER_mlp_decoder}, graph neural networks~\cite{PLACEHOLDER_gnn_decoder}, recurrent architectures~\cite{PLACEHOLDER_rnn_decoder}, and transformers~\cite{PLACEHOLDER_transformer_decoder}. These works focus primarily on decoder \emph{accuracy}---matching or exceeding MWPM---rather than on \emph{training methodology}. Our contribution is orthogonal: ExactK addresses a training-time confound (K-leakage) that affects any neural decoder whose internal representation can encode syndrome count, regardless of architecture. We do not claim that the factor-graph decoder used in this study outperforms MWPM at the final configuration; this comparison was not performed.

\paragraph{Factor-graph and BP-inspired decoders.} Belief-propagation decoders and their neural extensions have been applied to classical LDPC codes~\cite{PLACEHOLDER_bp_neural} and, more recently, to quantum codes~\cite{PLACEHOLDER_bp_quantum}. Our bipartite factor-graph architecture differs in that it preserves the DEM's $k{>}2$ hyperedge structure without clique expansion, and in that its readout pools only error nodes associated with the logical observable.

\paragraph{Shortcut learning and debiasing.} The phenomenon of models exploiting spurious correlations is well documented in the broader ML literature~\cite{PLACEHOLDER_shortcut_learning}. Group distributionally robust optimization (DRO)~\cite{PLACEHOLDER_group_dro} addresses spurious correlations by reweighting or resampling groups. ExactK shares the philosophy of within-group comparison but differs in a critical respect: in QEC, the ``group variable''~$K$ is \emph{physically meaningful}, not purely spurious. Global debiasing methods that treat $K$ as a nuisance variable destroy legitimate signal (Sec.~\ref{sec:prior_mitigation}). ExactK's design---constraining only the auxiliary ranking loss to same-$K$ pairs while leaving the main loss unrestricted---is specific to this partially-legitimate-correlation structure.

\paragraph{Pairwise ranking losses.} Hinge-based ranking losses have a long history in learning-to-rank~\cite{PLACEHOLDER_ranknet} and metric learning~\cite{PLACEHOLDER_contrastive}. ExactK applies constrained pair mining (same-$K$, opposite-label) to a debiasing objective rather than a ranking-quality objective. The constraint $\Delta K{=}0$ is the key distinguishing feature; without it, the mechanism degrades to a form of global decorrelation that fails for the reasons described in Sec.~\ref{sec:prior_mitigation}.
